Convergence rate of MAP estimates \\ for the exponential family

### 2021-01-26 main.tex
Add Cramer-Rao discussion

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
%\usepackage{subfigure}
\usepackage{subcaption}
\usepackage{hyperref}       % hyperlinks
\usepackage{tikz}           % define for loops

@@

In the limit, the MAP estimates reach the Cramer-Rao lower bound. This can be seen by approximating the Bregman divergence using a second order Taylor expansion
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    &= \half (\mu - \mu^*)^\top \nabla^2\conj(\mu^*)(\mu - \mu^*)  
    + O(\norm{\mu - \mu^*}^3)\\
    &=\half  \norm{\mu^* - \mu}^2_{\nabla^2\conj(\mu^*)}
    + O(\norm{\mu - \mu^*}^3)
\end{align}
where we introduced the Mahalanobis distance induced by the matrix 
\begin{align}
    \nabla^2\conj(\mu^*) 
    = \nabla^2\logpart(\natp^*)^{-1} 
    = \Cov_{\natp^*}[T(X)]^{-1} 
    =: \mSigma^{-1}  \; .
\end{align}
To exploit this approximation, let us extend the $\ell^2$ norm results to a Mahalanobis distance induced by matrix $\mM$
\begin{align}
	\E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}_\mM^2 = \inv{n} \Tr(\mM \mSigma)
\end{align}
where $\mSigma$ is the covariance of $T(X)$. We retrieve $\nicefrac{\sigma^2}{n}$, the variance divided by the number of samples, when the metric is the identity $\mM=\mI$.  For the MLE we get
\begin{align}
	\E \bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right ) = \frac{d}{2n} + O(n^{- \frac{3}{2}})
\end{align}
\RLP{Maybe that is Cramer-Rao, but check again.} And for the MAP we  get 
\begin{align}
	\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2_{\mSigma^{-1}}
	&= \frac{n d}{(n+n_0)^2}  +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} \\
	&= O\left(\frac{d}{n} \right) + O\left(\frac{\norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} }{n^2} \right) \; .
\end{align}
Remark how the variance does not even appear, only the dimension divided by $n$ really matters. I find this result quite spectacular : the convergence speed of MLE is not affected by the covariance of sufficient statistics, and for MAP it matters only in the $O(n^{-2})$ term.
Anyway that's all good for asymptotic results, but we are interested in a finite sample analysis. How do we get this ?

@@

Given this posterior, a natural quantity to consider is the expected loss. Thus we might ask "what is the expectation of the posterior expected suboptimality?"
\begin{align}
    \expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
\end{align}

@@

\paragraph{Prior} A conjugate prior over the natural parameter $\natp$ is 
\begin{align}
	-\log \pi(\natp)
	&= n_0 \bregman(\natp ; \natp_0) + \cst \\
	&= n_0 \logpart(\natp) - n_0\mu_0 \natp + \cst \\
	&= -\half[n_0] \log(-\natp) + n_0\mu_0 (-\natp) + \cst \; . \\
	\pi (\natp) &= (-\natp)^{1 + \half[n_0] -1 } e^{-n_0 \mu_0 (-\natp)} / Z
\end{align}
In words, $-\natp$ is a random variable from the exponential family with sufficient statistics $(-\natp, \log(-\natp))$ and with natural parameters $(-n_0 \mu_0, \half[n_0])$. This is the definition of a gamma distribution with shape parameter $1 + \half[n_0]$ and with rate parameter $n_0 \mu_0$.

>>>>

\usepackage[top=2cm, bottom=2cm]{geometry}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
%\usepackage{subfigure}
\usepackage{subcaption}
\usepackage{hyperref}       % hyperlinks
\usepackage{tikz}           % define for loops

@@

In the limit, the MAP estimates reach a rate $O(1/2n)$. This is shown by approximating the Bregman divergence using a second order Taylor expansion
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    &= \half (\mu - \mu^*)^\top \nabla^2\conj(\mu^*)(\mu - \mu^*)  
    + O(\norm{\mu - \mu^*}^3)\\
    &=\half  \norm{\mu^* - \mu}^2_{\nabla^2\conj(\mu^*)}
    + O(\norm{\mu - \mu^*}^3)
\end{align}
where we introduced the Mahalanobis distance induced by the matrix 
\begin{align}
    \nabla^2\conj(\mu^*) 
    = \nabla^2\logpart(\natp^*)^{-1} 
    = \Cov_{\natp^*}[T(X)]^{-1} 
    =: \mSigma^{-1}  \; .
\end{align}
To exploit this approximation, let us extend the $\ell^2$ norm results to a Mahalanobis distance induced by matrix $\mM$
\begin{align}
	\E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}_\mM^2 = \inv{n} \Tr(\mM \mSigma)
\end{align}
where $\mSigma$ is the covariance of $T(X)$. We retrieve $\nicefrac{\sigma^2}{n}$, the variance divided by the number of samples, when the metric is the identity $\mM=\mI$.  For the MLE we get
\begin{align}
	\E \bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right ) = \frac{d}{2n} + O(n^{- \frac{3}{2}})
\end{align}
and for the MAP we  get 
\begin{align}
	\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2_{\mSigma^{-1}}
	&= \frac{n d}{(n+n_0)^2}  +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} \\
	&= O\left(\frac{d}{n} \right) + O\left(\frac{\norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} }{n^2} \right) \; .
\end{align}
Remark how the variance does not even appear, only the dimension divided by $n$ really matters. I find this result quite spectacular : the convergence speed of MLE is not affected by the covariance of sufficient statistics, and for MAP it matters only in the $O(n^{-2})$ term.
Indeed this is because it is hidden within the Bregman divergence loss. 
Anyway that's all good for asymptotic results, but we are interested in a finite sample analysis. How do we get this ?

\subsection{Cramer-Rao Lower Bound}
In statistics, there is this famous theorem known as the Cramer-Rao lower bound. In its simplest form, it states that the covariance of an unbiased estimator of a distribution's parameter $\natp$ is lower bounded by the inverse Fisher information matrix, divided by $n$
\begin{align}
    \Cov(\hat \natp_n)  
    &\succeq \inv{n} \cI(\natp_*)^{-1} \\
    \iff
    \expect[\cD\sim \natp_*]{(\hat \natp_n - \natp_*)(\hat \natp_n - \natp_*)^T}
    &\succeq \inv{n} \expect[X\sim \natp_*]{-\nabla^2 \log p(X|\natp_*)}^{-1} \; .
\end{align}
This is a typical statistic bound, which is concerned with the $\ell^2$ accuracy of an estimator, via its covariance. In contrast, here we are concerned with the quality of the estimator in terms of KL-divergence, eg "how accurately am I modeling this distribution?"

@@

Given this posterior, a natural quantity to consider is the expected loss. Thus we might ask "what is the expectation of the posterior expected suboptimality?"
\begin{align}
    \expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
\end{align}
This quantity is even harder to estimate than the expected MAP suboptimality, because of the double integral. We can get a closed form for the age-old $\ell^2$ loss of $\cN(\mu, 1)$, which yields the exact rate $\inv{n}$. This is exactly twice the expected loss of the MAP.

@@

\paragraph{Prior} A conjugate prior over the natural parameter $\natp$ is 
\begin{align}
	-\log \pi(\natp)
	&= n_0 \bregman(\natp ; \natp_0) + \cst \\
	&= n_0 \logpart(\natp) - n_0\mu_0 \natp + \cst \\
	&= -\half[n_0] \log(-\natp) + n_0\mu_0 (-\natp) + \cst \; . \\
	\pi (\natp) &= (-\natp)^{1 + \half[n_0] -1 } e^{-n_0 \mu_0 (-\natp)} / Z
\end{align}
In words, $-\natp$ is a random variable from the exponential family with sufficient statistics $(-\natp, \log(-\natp))$ and with natural parameters $(-n_0 \mu_0, \half[n_0])$. This is the definition of a Gamma distribution with shape parameter $1 + \half[n_0]$ and with rate parameter $n_0 \mu_0$.
